Revisiting the Relationship between Short-Chain Fatty Acids, the Microbiota, and Colorectal Tumors

### 2018-08-08 submission/manuscript.Rmd
fix figure legend captions

The authors thank the Great Lakes-New England Early Detection Research Network for providing the fecal samples that were used in this study. We would also like to thank Kwi Kim and Thomas M Schmidt for their help in running the short-chain fatty acid analysis on the High-Performance Liquid Chromatography machine at the University of Michigan. Salary support for Marc A. Sze came from the Canadian Institute of Health Research and NIH grant UL1TR002240. Salary support for Patrick D. Schloss came from NIH grants P30DK034933 and 1R01CA215574.

@@

**Figure 1. No change in SCFA measurements was observed between normal, adenoma, and carcinoma individuals using HPLC.** Acetate concentrations in fecal samples of individuals without colon tumors, adenomas, and carcinomas (A). Butyrate concentrations in fecal samples of individuals without colon tumors, adenomas, and carcinomas (B). Propionate concentrations in fecal samples of individuals without colon tumors, adenomas, and carcinomas (C). The black lines indicate the median SCFA concentration. Acetate concentrations in fecal samples before and after treatment for adenoma (yellow) and carcinoma (red) (D). Butyrate concentrations in fecal samples before and after treatment for adenoma (yellow) and carcinoma (red) (E). Propionate concentrations in fecal samples before and after treatment for adenoma (yellow) and carcinoma (red) (F). The black dots and lines represent the median change in SCFA concentration.

**Figure 2. No change in butyrate producing genes identified between normal, adenoma, and carcinoma individuals.** Imputed gene relative abundance of important butyrate pathway genes using PICRUSt (A). Counts per million (corrected for size and number of contigs in an OPF) for the Butyrate Kinase gene (B). The other butyrate pathway genes from the PICRUSt analysis did not align to any of the OPFs in the metagenome analysis.

**Figure 3. Patients with adenomas had the higest number of significant negative correlations between OTU relative abundance and SCFA concentration.** Colors denote the family or lowest taxonomic ID that an OTU classified to. Fewer significant positive correlations were observed overall. Additionally, the differences in the number of significant positive correlations between patients with adenomas versus individuals without tumors (normal) and patients with carcinomas was not as pronounced as the number of significant negative correlations.

**Figure 4. SCFA concentrations do not improve OTU-based Random Forest models.** The area under the curve of 100 different 80/20 split OTU-based normal versus adenoma 10-fold CV models with and without SCFAs (A). The top 10 most important OTUs or SCFAs in the SCFA and OTU adenoma model (B). The top 10 most important OTUs in the OTU adenoma model (C). The area under the curve of 100 different 80/20 OTU-based normal versus carcinoma 10-fold CV models with and without SCFAs (D). The top 10 most important OTUs or SCFAs in the SCFA and OTU carcinoma model (E). The top 10 most important OTUs in the OTU carcinoma model (F). For (A) and (D) the black line represents the median AUC. The dotted line highlights an AUC of 0.5.

**Figure 5. OTU-based regression Random Forest models of SCFA concentrations.** The train and test correlation between actual and predicted values from 100 different 80/20 split OTU-based models with 10-fold CV using regression Random Forest (A). The model accuracy of predicted SCFA concentrations differed between individuals without tumors, patients with adenomas, and patients with carcinomas. Generally, patients with carcinomas had predicted concentrations closest to their actual measured concentration (B). The top 10 OTUs based on mean decrease in accuracy (MDA) for each SCFA model, colored by their lowest taxonomic identification (C).

@@

**Figure S1. Patients with adenomas had the higest number of significant differences in OTU relative abundance between high/low SCFA groups.** Colors denote the family or lowest taxonomic ID that an OTU classified to. Fewer significant OTUs were observed in individuals without tumors (normal) and patients with carcinomas versus patients with adenomas.

**Figure S2. OTU-based classification Random Forest models of high/low SCFA groups based on overall SCFA median concentration.** The train and test results of 100 different 80/20 split OTU-based models with 10-fold CV based on higher or lower than the median SCFA concentration using classification Random Forest (A). The model accuracy of predicted high/low SCFA groups differed between individuals without tumors, patients with adenomas, and patients with carcinomas. Patients with adenomas consistently had the best classification accuracy (B). The top 10 OTUs based on mean decrease in accuracy (MDA) for each SCFA model, colored by their lowest taxonomic identification (C).

>>>>

The authors thank the Great Lakes-New England Early Detection Research Network for providing the fecal samples that were used in this study. We would also like to thank Kwi Kim and Thomas M Schmidt for their help in running the short-chain fatty acid analysis on the High-Performance Liquid Chromatography machine at the University of Michigan. We would also like to thank Ada Hagan for providing valuable feedback on earlier drafts of the manuscript. Salary support for Marc A. Sze came from the Canadian Institute of Health Research and NIH grant UL1TR002240. Salary support for Patrick D. Schloss came from NIH grants P30DK034933 and 1R01CA215574.

@@

**Figure 1. No change in SCFA measurements was observed between normal, adenoma, and carcinoma individuals using HPLC.** Acetate (A), butyrate (B), and propionate (C) concentrations in fecal samples of individuals without colon tumors (Control), adenomas, and carcinomas (C). The black lines indicate the median SCFA concentration. Acetate (D), butyrate (E), and propionate (F) concentrations in fecal samples before and after treatment for adenoma (yellow) and carcinoma (red). The black dots and lines represent the median change in SCFA concentration.

**Figure 2. No change in the butyrate kinase gene abudnance was observed between normal, adenoma, and carcinoma individuals.** Imputed gene relative abundance of important butyrate pathway genes using PICRUSt (A). Counts per million (corrected for size and number of contigs in an OPF) for the Butyrate Kinase gene (B).

**Figure 3. Patients with adenomas had the higest number of statistically significant correlations between OTU relative abundance and SCFA concentration.** Colors denote the family or lowest taxonomic ID that an OTU classified to. Fewer significant positive correlations were observed overall. Additionally, the differences in the number of significant positive correlations between patients with adenomas versus individuals without tumors (normal) and patients with carcinomas was not as pronounced as the number of significant negative correlations.

**Figure 4. SCFA concentrations do not improve OTU-based Random Forest models.** Individuals without tumors versus patients with adenomas model with and without SCFAs area under the curve (AUC) from 100 80/20 splits with 10-fold cross-validation (A). The top 10 most important OTUs or SCFAs in the SCFA and OTU adenoma model (B). The top 10 most important OTUs in the OTU adenoma model (C). Individuals without tumors versus patients with carcinomas model with and without SCFAs area under the curve (AUC) from 100 80/20 splits with 10-fold cross-validation (D). The top 10 most important OTUs or SCFAs in the SCFA and OTU carcinoma model (E). The top 10 most important OTUs in the OTU carcinoma model (F). For (A) and (D) the black line represents the median AUC. The dotted line highlights an AUC of 0.5.

**Figure 5. OTU-based regression Random Forest models of SCFA concentrations.** The train and test correlation between actual and predicted values from 100 different 80/20 split OTU-based models with 10-fold CV using regression Random Forest (A). The model accuracy of predicted SCFA concentrations differed between individuals without tumors, patients with adenomas, and patients with carcinomas. Differences between the actual and predicted SCFA concentrations based on whether individuals did not have tumors (Normal), had an adenoma, or had a carcinoma (B). The top 10 OTUs based on mean decrease in accuracy (MDA) for each SCFA model, colored by their lowest taxonomic identification (C).

@@

**Figure S1. Patients with adenomas had the higest number of significant differences in OTU relative abundance between high/low SCFA groups.** Colors denote the family or lowest taxonomic ID that an OTU classified to.

**Figure S2. OTU-based classification Random Forest models of high/low SCFA groups based on overall SCFA median concentration.** The train and test results of 100 different 80/20 split OTU-based models with 10-fold CV based on higher or lower than the median SCFA concentration using classification Random Forest (A). The model accuracy of predicted high/low SCFA groups differed between individuals without tumors, patients with adenomas, and patients with carcinomas (B). The top 10 OTUs based on mean decrease in accuracy for each SCFA model, colored by their lowest taxonomic identification (C).

### 2018-08-08 submission/manuscript.tex
fix figure legend captions

\textbf{Figure 1. No change in SCFA measurements was observed between
normal, adenoma, and carcinoma individuals using HPLC.} Acetate
concentrations in fecal samples of individuals without colon tumors,
adenomas, and carcinomas (A). Butyrate concentrations in fecal samples
of individuals without colon tumors, adenomas, and carcinomas (B).
Propionate concentrations in fecal samples of individuals without colon
tumors, adenomas, and carcinomas (C). The black lines indicate the
median SCFA concentration. Acetate concentrations in fecal samples
before and after treatment for adenoma (yellow) and carcinoma (red) (D).
Butyrate concentrations in fecal samples before and after treatment for
adenoma (yellow) and carcinoma (red) (E). Propionate concentrations in
fecal samples before and after treatment for adenoma (yellow) and
carcinoma (red) (F). The black dots and lines represent the median
change in SCFA concentration.

\textbf{Figure 2. No change in butyrate producing genes identified
between normal, adenoma, and carcinoma individuals.} Imputed gene
relative abundance of important butyrate pathway genes using PICRUSt
(A). Counts per million (corrected for size and number of contigs in an
OPF) for the Butyrate Kinase gene (B). The other butyrate pathway genes
from the PICRUSt analysis did not align to any of the OPFs in the
metagenome analysis.

\textbf{Figure 3. Patients with adenomas had the higest number of
significant negative correlations between OTU relative abundance and
SCFA concentration.} Colors denote the family or lowest taxonomic ID
that an OTU classified to. Fewer significant positive correlations were
observed overall. Additionally, the differences in the number of
significant positive correlations between patients with adenomas versus
individuals without tumors (normal) and patients with carcinomas was not
as pronounced as the number of significant negative correlations.

\textbf{Figure 4. SCFA concentrations do not improve OTU-based Random
Forest models.} The area under the curve of 100 different 80/20 split
OTU-based normal versus adenoma 10-fold CV models with and without SCFAs
(A). The top 10 most important OTUs or SCFAs in the SCFA and OTU adenoma
model (B). The top 10 most important OTUs in the OTU adenoma model (C).
The area under the curve of 100 different 80/20 OTU-based normal versus
carcinoma 10-fold CV models with and without SCFAs (D). The top 10 most
important OTUs or SCFAs in the SCFA and OTU carcinoma model (E). The top
10 most important OTUs in the OTU carcinoma model (F). For (A) and (D)
the black line represents the median AUC. The dotted line highlights an
AUC of 0.5.

\textbf{Figure 5. OTU-based regression Random Forest models of SCFA
concentrations.} The train and test correlation between actual and
predicted values from 100 different 80/20 split OTU-based models with
10-fold CV using regression Random Forest (A). The model accuracy of
predicted SCFA concentrations differed between individuals without
tumors, patients with adenomas, and patients with carcinomas. Generally,
patients with carcinomas had predicted concentrations closest to their
actual measured concentration (B). The top 10 OTUs based on mean
decrease in accuracy (MDA) for each SCFA model, colored by their lowest
taxonomic identification (C).

@@

\textbf{Figure S1. Patients with adenomas had the higest number of
significant differences in OTU relative abundance between high/low SCFA
groups.} Colors denote the family or lowest taxonomic ID that an OTU
classified to. Fewer significant OTUs were observed in individuals
without tumors (normal) and patients with carcinomas versus patients
with adenomas.

\textbf{Figure S2. OTU-based classification Random Forest models of
high/low SCFA groups based on overall SCFA median concentration.} The
train and test results of 100 different 80/20 split OTU-based models
with 10-fold CV based on higher or lower than the median SCFA
concentration using classification Random Forest (A). The model accuracy
of predicted high/low SCFA groups differed between individuals without
tumors, patients with adenomas, and patients with carcinomas. Patients
with adenomas consistently had the best classification accuracy (B). The
top 10 OTUs based on mean decrease in accuracy (MDA) for each SCFA
model, colored by their lowest taxonomic identification (C).

>>>>

\textbf{Figure 1. No change in SCFA measurements was observed between
normal, adenoma, and carcinoma individuals using HPLC.} Acetate (A),
butyrate (B), and propionate (C) concentrations in fecal samples of
individuals without colon tumors (Control), adenomas, and carcinomas
(C). The black lines indicate the median SCFA concentration. Acetate
(D), butyrate (E), and propionate (F) concentrations in fecal samples
before and after treatment for adenoma (yellow) and carcinoma (red). The
black dots and lines represent the median change in SCFA concentration.

\textbf{Figure 2. No change in the butyrate kinase gene abudnance was
observed between normal, adenoma, and carcinoma individuals.} Imputed
gene relative abundance of important butyrate pathway genes using
PICRUSt (A). Counts per million (corrected for size and number of
contigs in an OPF) for the Butyrate Kinase gene (B).

\textbf{Figure 3. Patients with adenomas had the higest number of
statistically significant correlations between OTU relative abundance
and SCFA concentration.} Colors denote the family or lowest taxonomic ID
that an OTU classified to. Fewer significant positive correlations were
observed overall. Additionally, the differences in the number of
significant positive correlations between patients with adenomas versus
individuals without tumors (normal) and patients with carcinomas was not
as pronounced as the number of significant negative correlations.

\textbf{Figure 4. SCFA concentrations do not improve OTU-based Random
Forest models.} Individuals without tumors versus patients with adenomas
model with and without SCFAs area under the curve (AUC) from 100 80/20
splits with 10-fold cross-validation (A). The top 10 most important OTUs
or SCFAs in the SCFA and OTU adenoma model (B). The top 10 most
important OTUs in the OTU adenoma model (C). Individuals without tumors
versus patients with carcinomas model with and without SCFAs area under
the curve (AUC) from 100 80/20 splits with 10-fold cross-validation (D).
The top 10 most important OTUs or SCFAs in the SCFA and OTU carcinoma
model (E). The top 10 most important OTUs in the OTU carcinoma model
(F). For (A) and (D) the black line represents the median AUC. The
dotted line highlights an AUC of 0.5.

\textbf{Figure 5. OTU-based regression Random Forest models of SCFA
concentrations.} The train and test correlation between actual and
predicted values from 100 different 80/20 split OTU-based models with
10-fold CV using regression Random Forest (A). The model accuracy of
predicted SCFA concentrations differed between individuals without
tumors, patients with adenomas, and patients with carcinomas.
Differences between the actual and predicted SCFA concentrations based
on whether individuals did not have tumors (Normal), had an adenoma, or
had a carcinoma (B). The top 10 OTUs based on mean decrease in accuracy
(MDA) for each SCFA model, colored by their lowest taxonomic
identification (C).

@@

\textbf{Figure S1. Patients with adenomas had the higest number of
significant differences in OTU relative abundance between high/low SCFA
groups.} Colors denote the family or lowest taxonomic ID that an OTU
classified to.

\textbf{Figure S2. OTU-based classification Random Forest models of
high/low SCFA groups based on overall SCFA median concentration.} The
train and test results of 100 different 80/20 split OTU-based models
with 10-fold CV based on higher or lower than the median SCFA
concentration using classification Random Forest (A). The model accuracy
of predicted high/low SCFA groups differed between individuals without
tumors, patients with adenomas, and patients with carcinomas (B). The
top 10 OTUs based on mean decrease in accuracy for each SCFA model,
colored by their lowest taxonomic identification (C).
